Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2021-02-03 paper/main.tex
paper

\subsubsection{Datasets}
%
%For the LOP, we generated synthetic instances as follows. First, we create a sample $S$ of 200 permutations of size $n=20$ from a MM with $\sigma_0$ equal to the identity permutation and $\theta=-\log(\textit{Uniformity})$, where  $\textit{Uniformity}\in\{0.5,$ $0.7,$ $0.9\}$, i.e., from peaked distributions to almost uniform distributions. Then, we generate matrix $a$ as $a_{i,j} = \sum_{\sigma\in S}\sum_{i<j}\mathbb{I}\{\sigma(i)<\sigma(j)\}$, where ${\mathbb{I}\colon S \times S \to \{0,1\}}$ is the indicator function. In principle, these should be best-case instances for UMM since their fitness landscape closely matches the Mallows model.%\MANUEL{Ekhine: Could you revise this text? I wanted to explain why we use synthetic datasets.}
We consider real LOP datasets from the LOLIB
repository.\footnote{\url{http://grafo.etsii.urjc.es/optsicom/lolib/}} In
particular, we have considered two instances of each class, focusing on relatively instances (the computation of CEGO grows rapidly with instance size):
\texttt{N-p40-01}, \texttt{N-p40-02}, \texttt{N-p50-01}, \texttt{N-p50-02} (RandB), 
lop/RandB/
lop/IO/N-t59d11xx
lop/IO/N-t59b11xx
lop/SGB/N-sgb75.02
lop/SGB/N-sgb75.01
RandB/
RandB/N-p50-02

: IO, Random instances of type A and
Random instances of type B of various sizes $n \in \{50, 100, 150, 200,
250\}$. These are harder problems than the synthetic instances, not only
because of the larger search space, but also because of the different instance
types.\MANUEL{Needs updating}

In the case of the PFSP, we consider the same instances as
\citet{ZaeStoBar2014:ppsn}, i.e., \texttt{reC05}, \texttt{reC13} and
\texttt{reC19} for the PFSP, with $n \in \{20, 20, 30\}$, respectively.

@@

The first part of the experimental analysis concerns the {analyses} of the learning rate, which corresponds to parameter $\rho$ of uBorda. As stated in Section~\ref{sec:sampling}, $\rho$ is automatically set at each iteration in such a way that the $100r_1\%$ of the samples have the $100r_2\%$ of the weight. In this regard, we tried a few different values of parameters $r_1$ and $r_2$ in all the instances. Only values of $r_1 < r_2$ make sense and $r_1 = r_2 = 0.5$ would produce equal weights in the learning step. Thus, we tried values $r_1 = \{0.1, 0.2, \dotsc, 0.5\}$ and $r_2 = \{0.6,\dotsc, 0.9, 0.99\}$. Figure~\ref{fig:heatmaps} shows the best solution found by each independent run (averaged over 10 runs) for instances: \texttt{N-t1d100.01} and \texttt{N-t1d150.01} for the LOP, \texttt{reC05} for the PFSP, \texttt{kra32} for the QAP. We terminated runs that took longer than 6 hours and those combinations are shown as white.  Results for all the datasets can be found in the supplemental material~(\supplement).\MANUEL{TODO: Update (mine)}

\begin{figure}[tb]
\centering%
\begin{minipage}{0.49\linewidth}
  \centering
    \texttt{N-t59d11xx} (LOP)\\[-0.2ex]
    \includegraphics[width=\textwidth]{../img/heatmap_lop_IO_N-t59d11xx}
  \end{minipage}
  \begin{minipage}{0.49\linewidth}
    \centering
    \texttt{N-sgb75.02} (LOP)\\[-0.2ex]
 \includegraphics[width=\textwidth]{../img/heatmap_lop_SGB_N-sgb75_02}
\end{minipage}\\\vskip 1em
\begin{minipage}{0.49\linewidth}
  \centering
  \texttt{reC13} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/heatmap_pfsp_rec13_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \centering
  \texttt{reC19} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/heatmap_pfsp_rec19_txt}
\end{minipage}
% \begin{minipage}{0.49\linewidth}
%   \centering
%   \texttt{kra32} (QAP)\\[-0.5ex]
% \includegraphics[width=\textwidth]{../img/heatmap_kra32_dat}
% \end{minipage}
\caption{Mean fitness (over 10 runs) of different configurations of $r_1$ and $r_2$ (darker is better).\label{fig:heatmaps}}
\end{figure}

@@

\subsection{Analysis of UMM and CEGO on LOP, QAP and PFSP}
To  understand the behavior of the algorithms, we record the fitness of
the solution evaluated at each step of each run and we average those values
over 10 runs with different random seeds on the same problem instance.  As an
example, each line in Fig.~\ref{fig:lop_synth} shows the mean fitness of the
\ekhine{solutions evaluated by either UMM or CEGO on one synthetic LOP instance}{ojo}. The
shaded area shows one standard deviation around the mean. Ideally, each new
evaluation will monotonically improve in fitness, since each solution evaluated
helps to refine the model and leads to a better solution being evaluated in the
next step.  However, it is possible that new solutions are worse
than their predecessors. %Nevertheless, in this example, there is a clear
% overall improvement as more evaluations are performed.
The minimum values
achieved in the plot up to a particular number of function evaluations gives an
estimation of the mean fitness of the best-so-far solution up to that point in
the run. The algorithms do not return the final solution shown but rather the
best one (minimum fitness) of each run.

@@

The situation is very similar for the QAP, where both algorithms show the same
behaviour as in the PFSP for all instances tested, except for
\texttt{kra} instances. As shown in Fig~\ref{fig:kra32}, for \texttt{kra32} both
algorithms are able to improve over the number of evaluations, although not
with ease. As in the synthetic LOP instances, we again observe that CEGO does
improve faster than UMM initially, yet CEGO appears to get stuck around
evaluation 250, and UMM overtakes it around evaluation 300. We believe the
reason is the same as before, that is, the CEGO model is not able to adequately
reflect the ruggedness of the actual landscape and, hence, the best solutions
found by the GA searching the model are not improved solutions for the actual
problem. Interestingly, there are well-known differences between
\texttt{kra30a}, \texttt{kra30b}, \texttt{kra32}, on one hand, and \texttt{nug12}, \texttt{nug30} and
\texttt{tho32}, on the other~\citep{AnsBriGou2002qap}. However, we cannot at
this moment point out the root cause of this different behavior.

\begin{figure}[tp]
  \centering%
  \includegraphics[width=\textwidth]{../img/fitness_real_qap_kra32_dat}
  \caption{Mean fitness  (and standard deviation)  of each solution evaluated by each algorithm on the QAP instance \texttt{kra32}.\label{fig:kra32}}
  \vspace*{-1em}
\end{figure}

@@

In the LOP, differences are mostly in favour of UMM, due to the time limit of
144 hours, which resulted in CEGO typically achieving less than 200
evaluations. In contrast, UMM was always able to reach 400 evaluations in less
than 12.5 hours. If UMM was allowed to reach the 144 hours limit, the differences
would be even stronger.
%
In the PFSP, the differences are minor and/or not statistically
significant, with the qualification that both algorithms appear to be
performing a random search. Similar observations can be made for the QAP, where
only on the smallest instance \texttt{nug12}, there is a significant difference
in favour of CEGO.

@@

\begin{figure}[tp]
  \centering%
  \includegraphics[width=\textwidth]{../img/fitness_real_qap_kra32_dat}
  \caption{Mean fitness  (and standard deviation)  of each solution evaluated by each algorithm on the QAP instance \texttt{kra32}.}
  \label{fig:rankorder}
\end{figure}

@@

\subsubsection{Reproducibility.} Source code, datasets and scripts necessary to reproduce the results are available at \supplement.
%  \vspace*{-1em}
%\begin{acks}
%% MANUEL: Blind, so no acknowledgments yet.
%\paragraph*{Acknowledgements.} Blind
% M.\@ L\'opez-Ib\'a\~nez is a ``Beatriz Galindo'' Senior Distinguished Researcher (BEAGAL 18/00053) funded by the Ministry of Science and Innovation of the Spanish Government.
%
%\manuel{}{: We should also thank the COST ACTION: Ask Carlos?} \ekhine{}{Sure!, esto no habra q ponerlo aun por ser blind, no?}
%Thanks to: Hao Wang (Leiden University) for pointing us to the arguments of
%\citep{EriPeaGar2019scalable}; Thomas Stützle (Université libre de Bruxelles) for suggesting works
%about differences between QAPLIB instances; and Ahmed Boujaada for the experimental support. 
  % supported by the \grantsponsor{GS501100001809}{National Natural
  %   Science Foundation of
  %   China}{http://dx.doi.org/10.13039/501100001809} under Grant
  % No.:~\grantnum{GS501100001809}{61273304}
  % and~\grantnum[http://www.nnsf.cn/youngscientists]{GS501100001809}{Young
  %   Scientists' Support Program}.
%\end{acks}

>>>>

\subsubsection{Datasets}
%
%For the LOP, we generated synthetic instances as follows. First, we create a sample $S$ of 200 permutations of size $n=20$ from a MM with $\sigma_0$ equal to the identity permutation and $\theta=-\log(\textit{Uniformity})$, where  $\textit{Uniformity}\in\{0.5,$ $0.7,$ $0.9\}$, i.e., from peaked distributions to almost uniform distributions. Then, we generate matrix $a$ as $a_{i,j} = \sum_{\sigma\in S}\sum_{i<j}\mathbb{I}\{\sigma(i)<\sigma(j)\}$, where ${\mathbb{I}\colon S \times S \to \{0,1\}}$ is the indicator function. In principle, these should be best-case instances for UMM since their fitness landscape closely matches the Mallows model.%\MANUEL{Ekhine: Could you revise this text? I wanted to explain why we use synthetic datasets.}
We consider real LOP datasets from the LOLIB
repository.\footnote{\url{http://grafo.etsii.urjc.es/optsicom/lolib/}} In
particular, we have taken two instances of various types focusing on relatively
small instances, since he computation time of CEGO grows rapidly with instance
size: \texttt{N-p40-01}, \texttt{N-p40-02}, \texttt{N-p50-01} and
\texttt{N-p50-02} of type \emph{RandB}; \texttt{N-t59d11xx} and
\texttt{N-t59b11xx} of type \emph{IO}; and \texttt{N-sgb75.01} and
\texttt{N-sgb75.02} of type \emph{SGB}.
% Random instances of type B of various sizes $n \in \{50, 100, 150, 200,
% 250\}$. These are harder problems than the synthetic instances, not only
% because of the larger search space, but also because of the different instance
% types.
In the case of the PFSP, we consider the same instances as
\citet{ZaeStoBar2014:ppsn}, i.e., \texttt{reC05}, \texttt{reC13},
\texttt{reC19} and \texttt{rec31}, with $n \in \{20, 20, 30, 50\}$ and $M \in \{5, 20, 10, 15\}$, respectively.

@@

The first part of the experimental analysis concerns the analysis of the learning rate, which corresponds to parameter $\rho$ of uBorda. As stated in Section~\ref{sec:sampling}, $\rho$ is automatically set at each iteration in such a way that the $100r_1\%$ of the samples have the $100r_2\%$ of the weight. In this regard, we tried a few different values of parameters $r_1$ and $r_2$ in all the instances. Only values of $r_1 < r_2$ make sense and $r_1 = r_2 = 0.5$ would produce equal weights in the learning step. Thus, we tried values $r_1 = \{0.1, 0.2, \dotsc, 0.5\}$ and $r_2 = \{0.6,\dotsc, 0.9, 0.99\}$. Figure~\ref{fig:heatmaps} shows the best solution found by each independent run (averaged over 10 runs) for instances: \texttt{N-t59d11xx} and \texttt{N-sgb75.02} for the LOP, and \texttt{reC13} and \texttt{reC19} for the PFSP. Results for all the datasets can be found in the supplemental material~(\supplement).

\begin{figure}[tb]
  \centering%
  \smaller
\begin{minipage}{0.49\linewidth}
  \centering
    \texttt{N-t59d11xx} (LOP)\\[-0.2ex]
    \includegraphics[width=\textwidth]{../img/heatmap_lop_IO_N-t59d11xx}
  \end{minipage}
  \begin{minipage}{0.49\linewidth}
    \centering
    \texttt{N-sgb75.02} (LOP)\\[-0.2ex]
 \includegraphics[width=\textwidth]{../img/heatmap_lop_SGB_N-sgb75_02}
\end{minipage}\\\vskip 1em
\begin{minipage}{0.49\linewidth}
  \centering
  \texttt{reC13} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/heatmap_pfsp_rec13_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \centering
  \texttt{reC19} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/heatmap_pfsp_rec19_txt}
\end{minipage}
% \begin{minipage}{0.49\linewidth}
%   \centering
%   \texttt{kra32} (QAP)\\[-0.5ex]
% \includegraphics[width=\textwidth]{../img/heatmap_kra32_dat}
% \end{minipage}
\caption{Mean fitness (over 10 runs) of different configurations of $r_1$ and $r_2$ (darker is better).\label{fig:heatmaps}}
\end{figure}

@@

\subsection{Analysis of UMM and CEGO on LOP and PFSP}
To  understand the behavior of the algorithms, we record the fitness of
the solution evaluated at each step of each run and we average those values
over 10 runs with different random seeds on the same problem instance.  As an
example, each line in Fig.~\ref{fig:lop_synth} shows the mean fitness of the
\ekhine{solutions evaluated by either UMM or CEGO on one synthetic LOP instance}{ojo}. The
shaded area shows one standard deviation around the mean. Ideally, each new
evaluation will monotonically improve in fitness, since each solution evaluated
helps to refine the model and leads to a better solution being evaluated in the
next step.  However, it is possible that new solutions are worse
than their predecessors. %Nevertheless, in this example, there is a clear
% overall improvement as more evaluations are performed.
The minimum values
achieved in the plot up to a particular number of function evaluations gives an
estimation of the mean fitness of the best-so-far solution up to that point in
the run. The algorithms do not return the final solution shown but rather the
best one (minimum fitness) of each run.

@@

% The situation is very similar for the QAP, where both algorithms show the same
% behaviour as in the PFSP for all instances tested, except for
% \texttt{kra} instances. As shown in Fig~\ref{fig:kra32}, for \texttt{kra32} both
% algorithms are able to improve over the number of evaluations, although not
% with ease. As in the synthetic LOP instances, we again observe that CEGO does
% improve faster than UMM initially, yet CEGO appears to get stuck around
% evaluation 250, and UMM overtakes it around evaluation 300. We believe the
% reason is the same as before, that is, the CEGO model is not able to adequately
% reflect the ruggedness of the actual landscape and, hence, the best solutions
% found by the GA searching the model are not improved solutions for the actual
% problem. Interestingly, there are well-known differences between
% \texttt{kra30a}, \texttt{kra30b}, \texttt{kra32}, on one hand, and \texttt{nug12}, \texttt{nug30} and
% \texttt{tho32}, on the other~\citep{AnsBriGou2002qap}. However, we cannot at
% this moment point out the root cause of this different behavior.

% \begin{figure}[tp]
%   \centering%
%   \includegraphics[width=\textwidth]{../img/fitness_real_qap_kra32_dat}
%   \caption{Mean fitness  (and standard deviation)  of each solution evaluated by each algorithm on the QAP instance \texttt{kra32}.\label{fig:kra32}}
%   \vspace*{-1em}
% \end{figure}

@@

In the LOP, differences are mostly in favour of UMM, due to the time limit of
144 hours, which resulted in CEGO typically achieving less than 200
evaluations. In contrast, UMM was always able to reach 400 evaluations in less
than 12.5 hours. If UMM was allowed to reach the 144 hours limit, the differences
would be even stronger.
%
In the PFSP, the differences are minor and/or not statistically
significant, with the qualification that both algorithms appear to be
performing a random search.

@@

@@

\subsubsection{Reproducibility} Source code, datasets and scripts necessary to reproduce the results are available at \supplement.
%  \vspace*{-1em}
%\begin{acks}
%% MANUEL: Blind, so no acknowledgments yet.
%\paragraph*{Acknowledgements.} Blind
% M.\@ L\'opez-Ib\'a\~nez is a ``Beatriz Galindo'' Senior Distinguished Researcher (BEAGAL 18/00053) funded by the Ministry of Science and Innovation of the Spanish Government.
%
%\manuel{}{: We should also thank the COST ACTION: Ask Carlos?} \ekhine{}{Sure!, esto no habra q ponerlo aun por ser blind, no?}
%Thanks to: Hao Wang (Leiden University) for pointing us to the arguments of
%\citep{EriPeaGar2019scalable}; Thomas Stützle (Université libre de Bruxelles) for suggesting works
%about differences between QAPLIB instances; and Ahmed Boujaada for the experimental support. 
  % supported by the \grantsponsor{GS501100001809}{National Natural
  %   Science Foundation of
  %   China}{http://dx.doi.org/10.13039/501100001809} under Grant
  % No.:~\grantnum{GS501100001809}{61273304}
  % and~\grantnum[http://www.nnsf.cn/youngscientists]{GS501100001809}{Young
  %   Scientists' Support Program}.
%\end{acks}
